### 2019-09-09 ic-journal/ccs.tex
Stage work revising the paper (2nd round)

We assume a possibly infinite set of names $\mathscr{L} = \{a, b,
\ldots\}$ forming input and $\overline{\mbox{output}}$ actions, plus a special invisible
action $\tau \notin \mathscr{L}$, and a set of variables $A, B,
\ldots$ for defining recursive behaviours.
The class of CCS processes is inductively
defined from $\nil$ \hl{(deadlock)} by the operators
of prefixing, parallel composition, summation (binary choice),
restriction, relabeling and recursion:
\begin{equation*}
\begin{array}{cccl}
\mu  & := & \tau\!\!\!\! & \midd \; a  \; \midd \;  \outC a  \\
P  & := & \nil\!\!\!\! & \midd \;  \mu . P \; \midd \;  P_1 |  P_2 \; \midd  \;
P_1 + P_2 \; \midd 
(\res a\!)\, P
\; \midd \; P\; [r\!f]  % relabelling
\;  \midd \;  A \; \midd \; \recu A  P
\end{array}
\end{equation*}
We sometimes omit a trailing $\nil$, e.g., writing $a|b$ for $a.\nil |b .\nil$.
The operational semantics of CCS is then given by means of
a Labeled Transition System (LTS) \hl{expressed in Structural Operational
Semantics (SOS) rules}, shown in Fig.~\ref{f:LTSCCS}.
The symmetric version of the two \hl{SOS} rules for
parallel composition and the rule for sum are omitted.
A CCS expression uses only \emph{weakly guarded sums} if all occurrences of
the sum operator are of the form $\mu_1.P_1 + \mu_2.P_2 + \ldots
+ \mu_n.P_n$, for some $n \geq 2$.
 The \emph{immediate derivatives} of a
process $P$ are the elements of the set $\{P' \st P \arr\mu P' \mbox{
  for some $\mu$}\}$.
We use $\ell$ to range over
visible actions (i.e.~inputs or outputs, excluding  $\tau$)
and $\alpha, \mu$  to range over all actions.
\begin{figure*}[t]
\begin{center}
\vskip .1cm
 $\displaystyle{  \over  \mu.  P    \arr\mu
P } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P + Q   \arr\mu
P'  } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P | Q   \arr\mu
P' | Q } $  $ \hb$   
\hskip .3cm
  $\; \;$  $\displaystyle{ P \arr{ a}P' \hk \hk  Q
\arr{\outC a }Q'  \over     P|  Q \arr{ \tau} P'
|  Q'  }$ 
\\
\vspace{.2cm}
$\displaystyle{ P \arr{\mu}P' \over
 (\res a\!)\, P   \arr{\mu} (\res a\!)\, P'} $ $ \mu \neq a, \outC a$
$ \hb$
%
$\displaystyle{ P \sub {\recu A P} A \arr{\mu}P' \over
 \recu A P   \arr{ \mu} P'  } $
\hskip .5cm  
$\displaystyle{ P \arr{\mu} P' \over
 P \;[r\!f] \arr{r\!f(\mu)} P' \;[r\!f]} $
$ \hb$ %  &
\end{center}
\caption{Structural Operational Semantics of CCS}
\label{f:LTSCCS}
\end{figure*}
Some standard notations for transitions: $\Arr\epsilon$ is the 
reflexive and transitive closure of $\arr\tau$, and 
$\Arr \mu $ is $\Arr\epsilon \arr\mu \Arr\epsilon$ (the
composition of the three relations).
Moreover,   
$ P \arcap \mu P'$ holds if $P \arr\mu P'$ or ($\mu =\tau$ and
$P=P'$); similarly 
$ P \Arcap \mu P'$ holds if $P \Arr\mu P'$ or ($\mu =\tau$ and
$P=P'$).
We write $P \:(\arr\mu)^n P'$ if $P$ can become $P'$ after performing
$n$ $\mu$-transitions. Finally, $P \arr\mu$ holds if there is $P'$
with $P \arr\mu P'$, and similarly for other forms of transitions.

\paragraph{Further notations}
We let $\R$, $\S$ range over binary relations, sometimes using 
 infix notation for them; e.g., 
$P \RR Q$ means that $(P,Q) \in \R$.
We use a tilde, as in  $\til P$, to denote tuples of processes with
countably many elements (thus the tuple may also be infinite.)
All notations are extended to tuples componentwise,
e.g., $\til P \RR \til Q$ means that $P_i \RR Q_i$, for each  
component $i$ of the tuples $\til P$ and $\til Q$.
And $\ct{\til P}$ is the process obtained by replacing each hole
$\holei i$ of the  context $\qct$ with $P_i$.  
We write $\ctx \R$ for the closure of a relation under contexts.
Thus $P\: \ctx \R\: Q$
means that there \hl{exists a} context $\qct$ and tuples $\til P,\til Q$ with
$P =  \ct{\til P}, Q =  \ct{\til Q}$ and $\til P \RR \til Q$.
We use the symbol $\DSdefi$ for abbreviations. For instance, $P \DSdefi G $, where
$G$ is some expression, means that $P$ stands for the expression $G$.
If $\leq$ is a preorder, then  $\geq$  is its inverse (and
conversely).

>>>>

\hl{Here we beliefly recall the core theory of CCS, up to the unique solution of
equation and contractions, with a focus on the syntactic differences with
other literature.}
We assume a possibly infinite set of names $\mathscr{L} = \{a, b,
\ldots\}$ forming input and $\overline{\mbox{output}}$ labels (actions), plus a special invisible
action $\tau \notin \mathscr{L}$, and \hl{another possibly infinite
  set of agent} variables $A, B,
\ldots \in \mathscr{X}$.
The class of CCS processes is then inductively
defined from $\nil$ \hl{(deadlock) and agent variables} by the operators
of \emph{prefixing} (.), \emph{parallel composition} ($|$),
\emph{summation}
(or \emph{binary choice}, $+$),
\emph{restriction} ($\nu$), \emph{relabeling} ($[\cdot]$) and
\emph{recursion} ($\texttt{rec}$):
\begin{equation*}
\begin{array}{cccl}
\mu  & := & \tau\!\!\!\! & \midd \; a  \; \midd \;  \outC a  \\
P  & := & \nil\!\!\!\! & \midd \;  \mu . P \; \midd \;  P_1 |  P_2 \; \midd  \;
P_1 + P_2 \; \midd 
(\res L)\, P
\; \midd \; P\; [\mathit{rf}]  % relabelling
\;  \midd \;  A \; \midd \; \recu A  P
\end{array}
\end{equation*}
\hl{Notice the (minor) differences with other CCS literature regarding the above definition:}
\begin{enumerate}
  \item Some literature adopts possibly infinite sums over an index
    set ($\sum_{i\in I} P_i$) while we only consider binary (thus
    finite) sums, which is less expressive.
  \item Our restriction operator takes a set of (input) labels $L
    \subseteq \mathscr{L}$, while some CCS literature only restricts
    one label (for each restriction operator);
  \item Our relabeling operator takes a relabeling function
    $\mathit{rf} \colon \mathscr{L} \cup \overline{\mathscr{L}} \cup
    \{\tau\} \rightarrow \mathscr{L} \cup \overline{\mathscr{L}} \cup
    \{\tau\}$, which
    is capable to handle multiple actions including $\tau$. A valid
    relabeling function $\mathit{rf}$ must satisfy $\mathit{rf}(\tau) =
    \tau$ and $\forall l\in\mathscr{L} \cup \overline{\mathscr{L}}.\, \mathit{rf}(\overline{l}) =
    \overline{\mathit{rf}(l)}$ (with $\bar{\bar l} = l$ for all $l \in
    \mathscr{L}$); Same expressiveness
    with single-label relabeling.
  \item The recursion operator $\texttt{rec}$ has the same expressiveness with
    \emph{process constants} adopted in some literature. For instance, a
    CCS process $a.C$ with the constant $C \DSdefi b.C$ can be
    expressed as $a.(\texttt{rec}\, C. (b.C))$.
\end{enumerate}
Besides, we sometimes omit \hl{the} trailing $\nil$, e.g., writing $a|b$ for $a.\nil |b .\nil$.
\hl{A CCS process $P$ may \emph{derive} to another one, say $P'$, under
  an action $\mu$, denoted by $P \arr{\mu} P'$.}
The \hl{transitional semantics} of CCS processes is given by means of
a Labeled Transition System (LTS) \hl{expressed in Structural Operational
Semantics (SOS) rules} shown in Fig.~\ref{f:LTSCCS}.
(The symmetric versions of the \hl{SOS} rules for
parallel composition and binary choice are omitted.)
A CCS processes uses only \emph{guarded sums} if all occurrences of
sums are in the form $a.P + b.Q$. \hl{(Otherwise we may call the sum
a \emph{direct} one, e.g. $P + Q$ if either $P$ or $Q$ are not in the
form of prefixes.)}
 The \emph{immediate derivatives} of a
process $P$ are elements of the set $\{P' \st P \arr\mu P' \mbox{
  for some $\mu$}\}$.
We use $\ell$ to range over
visible actions (i.e.~inputs or outputs, excluding  $\tau$)
and $\alpha, \mu$  to range over all actions.
\begin{figure*}[t]
\begin{center}
\vskip .1cm
 $\displaystyle{  \over  \mu.  P \arr\mu
P } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P + Q   \arr\mu
P'  } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P | Q   \arr\mu
P' | Q } $  $ \hb$   
\hskip .3cm
  $\; \;$  $\displaystyle{ P \arr{ a}P' \hk \hk  Q
\arr{\outC a }Q'  \over  P|  Q \arr{ \tau} P'
|  Q'  }$ 
\\
\vspace{.2cm}
$\displaystyle{ P \arr{\mu}P' \over
 (\res L\!)\, P \arr{\mu} (\res L\!)\, P'} $ $ \mu, \outC\mu \not\in L$
$ \hb$
%
$\displaystyle{ P \sub {\recu A P} A \arr{\mu}P' \over
 \recu A P   \arr{ \mu} P'  } $
\hskip .5cm  
$\displaystyle{ P \arr{\mu} P' \over
 P \;[r\!f] \arr{r\!f(\mu)} P' \;[r\!f]} $
$ \hb$ %  &
\end{center}
\caption{Structural Operational Semantics of CCS}
\label{f:LTSCCS}
\end{figure*}

Some standard notations for transitions: $\Arr\epsilon$ is the 
reflexive and transitive closure of $\arr\tau$, and 
$\Arr \mu $ is $\Arr\epsilon \arr\mu \Arr\epsilon$ (the
composition of the three relations).
Moreover,   
$P \arcap \mu P'$ holds if $P \arr\mu P'$ or $\mu =\tau \wedge P = P'$; similarly 
$P \Arcap \mu P'$ holds if $P \Arr\mu P'$ or $\mu =\tau \wedge P = P'$).
We write $P \:(\arr\mu)^n P'$ if $P$ can become $P'$ after performing
$n$ $\mu$-transitions. Finally, $P \arr\mu$ holds if there is $P'$
with $P \arr\mu P'$, and similarly for other forms of transitions.

\paragraph{Further notations}
We let $\R$, $\S$ range over binary relations, sometimes using 
 infix notation for them; e.g., 
$P \,\R\, Q$ means $(P,Q) \in \R$.
We use a tilde, as in  $\til P$, to denote tuples of processes with
countably many elements (thus the tuple may also be infinite.)
All notations are extended to tuples componentwise,
e.g., $\til P \,\R\, \til Q$ means $P_i \,\R\, Q_i$ for each  
index $i$ of the tuples $\til P$ and $\til Q$.
We use the symbol $\DSdefi$ for abbreviations. For instance, $P \DSdefi G $, where
$G$ is some expression, means that $P$ stands for the expression $G$.
If $\leq$ is a preorder, then  $\geq$  is its inverse (and
conversely).

% \paragraph{CCS contexts and equations}
% And $\ct{\til P}$ is the process obtained by replacing each hole
% $\holei i$ of the  context $\qct$ with $P_i$.  
% We write $\ctx \R$ for the closure of a relation under contexts.
% Thus $P\: \ctx \R\: Q$
% means that there \hl{exists a} context $\qct$ and tuples $\til P,\til Q$ with
% $P =  \ct{\til P}, Q =  \ct{\til Q}$ and $\til P \RR \til Q$.

### 2019-09-09 ic-journal/equa.tex
Stage work revising the paper (2nd round)

@@

Uniqueness of  solutions of equations \cite{Mil89} intuitively says that if  a context $\qct$ obeys
certain  conditions, 
then all processes $P$  that satisfy the equation $ P \wb \ct P$ are
bisimilar with each other.
We need variables to write equations. We  use
 capital
letters  $X,Y,Z$
 for  these variables and call them \emph{\behav\  variables}.
 The body of an equation is a CCS expression
possibly containing \behav\  variables. Thus such expressions, ranged
over by $E$, live in the CCS
grammar extended with \behav\  variables.
% , which we  call 
%  \emph{extended CCS}.

>>>>

\hl{In the CCS syntax given in previous section, there is the syntactic
possibility that an agent variable $A$ occurs
outside the recursion operator of the same variable (or, equivalently, a
process constant $C$ is used but undefined). We say such a variable is
\emph{free}, while those variables wrapped by their recursions are
\emph{bound}.}
\hl{A CCS \emph{tem} (or \emph{expression}, i.e. those with valid syntax)
without any free variable is called a CCS \emph{process}.}
\hl{For instance, the variable $X$ is free in $a.X + \recu Y (b.Y)$, while $Y$
is bound. There is also the syntactic possibility that the same
variable $X$ is
both free and bound, e.g. in $a.X + \recu X (b.X)$.
This does not cause any ambiguity, but sometimes leads to more
complex proofs.} (c.f. Section~\ref{sec:multivariate} for more details.)

\hl{In this paper (and the formalisation work), we reuse the free agent
variables as \emph{equation variables}. This eliminates the need of
another type for CCS equations, and we can also reuse the existing
variable substitution operation} (c.f. SOS rule for the Recursion in
Fig.~\ref{f:LTSCCS}) \hl{for the substitution of equation variables.
For example, the result of substituting equation variable $X$ to $\nil$ in $a.X +
\recu X (b.X)$, denoted by $(a.X + \recu X (b.X)) \sub {\nil} X$, is
$a.\nil + \recu X (b.X)$ with the part $\recu X (b.X)$
untouched. Multivariate substitutions are written in the same syntax,
e.g. $E \sub {\til P} {\til X}$. Whenever $\til X$ is clear from the
context, we may also write $E[\til P]$ instead of $E \sub {\til P} {\til
  X}$. ($E[P]$ for $E \sub P X$ in the case of a single variable $X$.)}

\hl{In fact, free agent variables have the same transitional behavior
as the deadlock $\nil$ if we strictly follow the existing SOS rules, as there is
no SOS rule for them at all. In fact, most CCS theorems still hold if
the involved CCS terms contain free variables. (The most notable
exceptions are all versions of unique solution of equations/contracts,
where all solutions must be \emph{pure} processes (without free variables).)}

@@

Milner's ``unique solution of equations'' theorems~\cite{Mil89} intuitively
say that, if a context $C$ \hl{(i.e., a CCS expression with possibly
free variables)} obeys certain conditions,
then all processes $P$ that satisfy the equation $P \wb \ct P$ are
bisimilar with each other.

% TODO

### 2019-09-09 ic-journal/expansion.tex
Stage work revising the paper (2nd round)

The equivalences we consider here are mainly \emph{weak} ones, in that they
abstract from the number of internal steps being performed:
\begin{definition}%[bisimilarity]
\label{d:wb}
A process relation ${\R}$ is a \emph{bisimulation} if, whenever
 $P\RR Q$\hl{,} % for all $\mu$ we have:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q \Arcap \mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$,implies that there is $P'$ such that $P \Arcap
  \mu P'$ and $P' \RR Q'$\hl{;}
\end{enumerate}
 $P$ and $Q$ are \emph{bisimilar},
written as $P \wb Q$, if $P \RR Q$ for some bisimulation $\R$.
\end{definition}

\hl{We sometimes call the bisimulation and bisimilarity defined above
the \emph{weak} bisimulation (and weak bisimilarity), to
distinguish it from the \emph{strong} bisimulation and strong bisimilarity} ($\sim$),
obtained by replacing the weak transition $Q\Arcap\mu Q'$ in the above
definition with the strong transition $Q \arr \mu Q'$ (same for the
other case).
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums). For this, Milner introduced \emph{observational congruence}, also called \emph{rooted
  bisimilarity} \cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \emph{rooted bisimilar}, written as $P
\rapprox Q$, \hl{if} % for all $\mu$:
%  for all $\mu\in \mathscr{L}\cup\{\tau\}$
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}

\hl{The above definition brings up a proof technique for proving rooted
bisimilarity results, starting from a given bisimulation.}
\begin{lemma}[Rooted bisimilarity by bisimulation]
\label{l:obsCongrByWeakBisim}
Given a (weak) bisimulation $\RR$, if two processes $P$ and $Q$
\hl{satisfy} the following properties \hl{(but not necessary $P\;\R\;Q$)}:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$ \hl{implies that there is} $P'$ such that $P
   \Arr\mu P'$ and $P' \RR Q'$.
\end{enumerate}
\hl{Then} $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\approx$ \cite{van2005characterisation}.
\end{theorem}

>>>>

The equivalences we consider here are mainly \emph{weak} ones, in that they
abstract from the number of internal steps being performed:
\begin{definition}%[bisimilarity]
\label{d:wb}
A process relation ${\R}$ is a \hl{(weak)} \emph{bisimulation} if, whenever
 $P\RR Q$\hl{,} % for all $\mu$ we have:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q \Arcap \mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$,implies that there is $P'$ such that $P \Arcap
  \mu P'$ and $P' \RR Q'$\hl{;}
\end{enumerate}
 $P$ and $Q$ are \hl{(weakly)} \emph{bisimilar},
written as $P \wb Q$, if $P \RR Q$ for some bisimulation $\R$.
\end{definition}

\hl{The \emph{strong} bisimulation and strong bisimilarity} ($\sim$)
can be obtained by replacing the weak transition $Q\Arcap\mu Q'$ in the above
definition with normal transitions $Q \arr \mu Q'$ (similar for the other case).
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums), \hl{i.e. from $P_1 \wb Q_1$ and $P_2 \wb Q_2$ one
  cannot obtain $P_1 + P_2 \wb Q_1 + Q_2$.}
For this, Milner introduced the concept of \emph{observational congruence}, also called \emph{rooted
  bisimilarity}~\cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \emph{rooted bisimilar}, written as $P
\rapprox Q$, \hl{if} % for all $\mu$:
%  for all $\mu\in \mathscr{L}\cup\{\tau\}$
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}

\hl{The above definition also brings up a proof technique for proving rooted
bisimilarity from a given bisimulation. The next
lemma is usually not mentioned in CCS literature but plays an
important role in proving some key results in this paper:}
\begin{lemma}[Rooted bisimilarity by bisimulation]
\label{l:obsCongrByWeakBisim}
Given a (weak) bisimulation $\RR$, if two processes $P$ and $Q$
\hl{satisfy} the following properties \hl{(but not necessary $P\;\R\;Q$)}:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$ \hl{implies that there is} $P'$ such that $P
   \Arr\mu P'$ and $P' \RR Q'$.
\end{enumerate}
\hl{Then} $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

\hl{One basic property of rooted bisimiarity is that, it is (like $\sim$)
a congruence (thus also an equivalence relation)
w.r.t.~all CCS operators except for the Recursion and is the
coarsest congruence contained in bisimilarity:}
(c.f. Section~\ref{s:coarsest} for more details)
\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\wb$.
\end{theorem}

### 2019-09-09 ic-journal/intro.tex
Stage work revising the paper (2nd round)

A prominent proof method for bisimulation, put forward by Robin Milner and widely used in his
landmark CCS book \cite{Mil89}, is the
\emph{unique solution of equations}, whereby two tuples of processes are
componentwise bisimilar if they are solutions 
of the same system of equations.
This method is important in verification techniques and tools
based on algebraic reasoning \cite{BaeBOOK,theoryAndPractice,RosUnder10}.

In Milner's versions of the unique-solution theorems for weak
equivalences,
 the  method has severe syntactical limitations:
the equations must be \emph{(strongly) guarded and sequential}. That is, the
variables of the equations may only be used underneath a visible
prefix and proceed, in the syntax tree, only by the sum and prefix operators.
One way of overcoming such limitations is to replace the equations
 with special inequations called
\emph{contractions} \cite{sangiorgi2015equations,sangiorgi2017equations}. Contraction is a
preorder that, roughly, places some efficiency
constraints on processes.  The uniqueness of solutions of a system of contractions
is defined as with systems of equations: any two solutions must be (componentwise) bisimilar.
The difference with equations is in the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem can be
removed.  One constraint that still remains in
\cite{sangiorgi2017equations} (in which the issue is bypassed using a more
restrictive CCS syntax)
is the occurrences of direct sums, due to the failure of the
substitutivity of contraction under direct sums.

The main goal of the work described in this paper is a %rather
comprehensive formalisation of the core theory of CCS in the HOL
theorem prover (HOL4),  with a focus on the theory of unique solutions
of equations and contractions.
The formalisation, however, is not confined to the theory of unique
solutions, but embraces a
significant portion the theory of CCS~\cite{Mil89}, mostly because the
theory of unique solutions relies on a large number of more fundamental results.
Indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g.~the fixed-point and substitutivity properties), the
basic properties of
rooted bisimilarity (the congruence induced by weak
bisimilarity, also called observation congruence), and
their algebraic laws. Further extensions (beyond Nesi
\cite{Nesi:1992ve}) include four versions of ``bisimulation up to''
techniques (e.g.~bisimulation up-to bisimilarity) \cite{Mil89,sangiorgi1992problem}, and the
expansion and contraction preorder (two
efficiency-like refinements of weak bisimilarity). Concerning rooted bisimilarity, the formalisation
includes \hl{Hennessy's Lemma} and \hl{Deng's Lemma},
 and two \hl{theorems} saying that rooted bisimilarity is the coarsest (largest)
 congruence contained in weak bisimilarity ($\wb$): one following Milner's
 book~\cite{Mil89}, with the hypothesis that no processes can use up
 all labels;
the other without such hypothesis, essentially formalising van Glabbeek's paper~\cite{van2005characterisation}.
Similar theorems are proved for the rooted contraction preorder.
In this respect, the work is \hl{also} an extensive experiment \hl{using} the HOL theorem prover with its
recent developments, including \hl{its coinductive relation-defining package}.

From the \hl{point of} view of CCS theory, this formalisation has offered us the possibility of
further refining the theory of unique solutions of contractions.
In particular, the existing \hl{theorem}~\cite{sangiorgi2017equations} has
placed limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.
We have thus refined the contraction-based proof technique, by moving to  
\emph{rooted contraction}, that is, the coarsest precongruence contained in the contraction
preorder. The resulting unique-solution theorem is now valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
constraints on the occurrences of sums.

Another benefit of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures.
%
% Examples are: the results that rooted bisimilarity and rooted contraction are,
% respectively, the coarsest congruence contained in weak bisimilarity 
% and the coarsest precongruence contained in the contraction preorder; 
% the result about unique solution(s) of equations for weak bisimilarity that uses the
% contraction preorder as an auxiliary relation, and other unique solution results (e.g.~%
% the one for rooted in which
% the auxiliary relation is rooted contraction); various forms of enhancements of the bisimulation
% proof method (the `up-to' techniques).
%
% rephrased: (TODO: still hard to read --Chun)
\hl{Such structural simililarities can be found, for instance, in the
following cases: the proofs that rooted bisimilarity and rooted
contraction are, respectively, the coarsest congruence contained in
weak bisimilarity and the coarsest precongruence contained in the
contraction preorder; the proofs about unique solution(s) of equations
for weak bisimilarity that uses the contraction preorder as an
auxiliary relation, and other unique solution results (e.g. the one
for rooted in which the auxiliary relation is rooted contraction); the
proofs about various forms of enhancements of the bisimulation proof
method (the `up-to' techniques).}
%
In these cases, moving between proofs there are only a few places in
which the HOL proof scripts have to be modified.
Then the successful termination of the proof gives us a guarantee that the proof is
complete and trustworthy, eliminating the risks 
of overlooking or missing details as in \emph{paper-and-pencil} proofs.

@@

Section~\ref{ss:ccs} presents basic background materials on CCS,
including its syntax, operational semantics, bisimilarity and rooted
bisimilarity.
Section~\ref{s:eq} discusses equations and contractions.
Section~\ref{ss:new} presents rooted contraction and the related
unique-solution result for rooted bisimilarity.
Section~\ref{s:for} highlights our formalisation in HOL4.
Finally,  Section~\ref{s:rel} and \ref{s:concl} discuss related work,
conclusions, and a few directions for future work.

>>>>

A prominent proof method for bisimulation, put forward by Robin Milner and widely used in his
landmark CCS book~\cite{Mil89}, is the
\emph{unique solution of equations}, whereby two tuples of processes are
componentwise bisimilar if they are solutions of the same system of equations.
This method is important in verification techniques and tools
based on algebraic reasoning~\cite{BaeBOOK,theoryAndPractice,RosUnder10}.

\hl{Milner's unique-solution theorem for weak bisimilarity, however,
has severe syntactic limitations:}
the equations must be both \emph{(strongly) guarded} and \emph{sequential}. That is,
variables of the equations \hl{can only occur underneath visible prefixes and summations.}
One way \hl{to overcome} such limitations is to replace the equations
with special inequations called
\emph{contractions}~\cite{sangiorgi2015equations,sangiorgi2017equations}. Contraction is a
preorder that, roughly, places some efficiency constraints on processes.
\hl{The unique solution of contractions is defined as with equations:}
any two solutions must be componentwise bisimilar.
The difference \hl{from} equations is \hl{at} the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem \hl{are
eliminated}. One constraint that still remains
in~\cite{sangiorgi2017equations}
(where the issue is bypassed using a more
restrictive CCS syntax)
is the occurrences of \emph{direct} sums (e.g.~$P + Q$) due to
the \hl{non-substitutivity of the contraction preorder in this case}.

The work \hl{behind} this paper is a
comprehensive formalisation of Milner's Calculus of Communicating
Systems (CCS) in the HOL theorem prover (HOL4),
with a focus towards the theory of unique solutions of equations and contractions.
\hl{Almost all results in Milner's CCS book}~\cite{Mil89} are covered, since
the unique-solution theorems rely on a large number of fundamental results.
Indeed the formalisation encompasses all basic properties of strong and weak
\hl{bisimilarities} (e.g.~the fixed-point and substitutivity properties), the
basic properties of rooted bisimilarity and their algebraic laws.
Further extensions include several versions of ``bisimulation up to''
techniques, the expansion and contraction preorder.
Concerning rooted bisimilarity, the formalisation
includes \hl{Hennessy's Lemma} and \hl{Deng's Lemma},
 and two \hl{theorems} saying that rooted bisimilarity is the coarsest (largest)
 congruence contained in weak bisimilarity ($\wb$): one is classical
 with the hypothesis that no process uses up all labels;
the other without such hypothesis, essentially formalised van Glabbeek's proof~\cite{van2005characterisation}.
Similar results are also proved for the rooted contraction preorder.
In this respect, the work is \hl{also} an extensive experiment \hl{using} the HOL theorem prover with its
recent developments, including \hl{its coinductive relation-defining package}.

From the \hl{point of} view of CCS theory, this formalisation has offered us the possibility of
further refining the theory of unique solutions of contractions.
In particular, Sangiorgi's result~\cite{sangiorgi2017equations} has
placed limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.
\hl{In this paper, the contraction-based proof technique is further
refined by moving from the contraction preorder} to
\emph{rooted contraction}, that is, the coarsest precongruence contained in the contraction
preorder. The resulting unique-solution theorem is now valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
constraints on the occurrences of sums.

Another benefit of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures.
% rephrased: (TODO: still hard to read --Chun)
\hl{Such structural simililarities can be found, for instance, in the
following cases: the proofs that rooted bisimilarity and rooted
contraction are, respectively, the coarsest congruence contained in
weak bisimilarity and the coarsest precongruence contained in the
contraction preorder; the proofs about unique solution(s) of equations
for weak bisimilarity that uses the contraction preorder as an
auxiliary relation, and other unique solution results (e.g. the one
for rooted in which the auxiliary relation is rooted contraction); the
proofs about various forms of enhancements of the bisimulation proof
method (the `up-to' techniques).}
%
In these cases, moving between proofs there are only a few places in
which the HOL proof scripts have to be modified.
Then the successful termination of the proof gives us a guarantee that the proof is
complete and trustworthy, eliminating the risks 
of overlooking or missing details as in \emph{paper-and-pencil} proofs.

\hl{The univariate versions of unique-solution theorems use
$\lambda$-functions to present univariate CCS equations,
while the multivariate versions require a
careful and detailed formalisation of CCS equations with equation
variables and multivariate variable substitutions.
In contrast to most literature, we have adopted the same
type for both CCS equations and processes: those undefined constants
in CCS terms are treated as equation (free) variables, and a CCS
\emph{process} is a CCS term without equation variables. As a result,
most CCS results except the unique-solution theorems are valid for
both CCS processes and equations.} (c.f. Section~\ref{sec:multivariate}
for more deails.)

@@

Section~\ref{ss:ccs} presents basic background materials on CCS,
including its syntax, operational semantics, bisimilarity and rooted
bisimilarity.
Section~\ref{s:eq} discusses equations and contractions.
Section~\ref{ss:new} presents rooted contraction and the related
unique-solution result for rooted bisimilarity.
Section~\ref{s:for} highlights our formalisation in HOL4.
In Section~\ref{sec:multivariate}, \hl{we describe the formalisation of
multivariate versions of unique-solution theorems.}
Finally,  Section~\ref{s:rel} and \ref{s:concl} discuss related work,
conclusions, and a few directions for future work.
